**Bayesian statistics for longitudinal studies in biomedical research**

### 2020-12-25 Manuscripts/Manuscript_AM_v1.Rmd
Changes in manuscript
Updated manuscript background section

A longitudinal study is defined as that which is designed to repeatedly measure a variable of interest in a group (or groups) of subjects. In biomedical research, this type of study arises when the investigator intends to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study). Clinical examples of this  approach in biomedical research include studies on breast and neck cancer [@sio2016;@kamstra2015]; in the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT). Longitudinal studies have used also to measure tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010]. From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach: it requires a lower number of subjects to reach a certain statistical power, and besides it being able to track the previously mentioned time-effect evolution on a group-by-group basis, it allows to determine the variability of the response within subjects [@guo2013; @fitzmaurice2012]. In other words,a longitudinal study permits to quantify how the variable changes within each subject across time.

Traditionally, a “frequentist” approach is used in biomedical research to derive inferences from the results of a longitudinal study. Such statistical view derives its name from the fact that it regards probability as a limiting frequency [wagenmakers2008] and its application is based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). This methodology makes two key assumptions regarding longitudinal data: a constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study  [@schober2018;@gueorguieva2004].

However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases [@ugrinowitsch2004], and the violation of this assumption increases the false positivity rate [@lane2016].
Moreover, it is unlikely that complete observations in all subjects are obtained in a biomedical study due to reasons that can be specific to different situations. In a clinical trial, voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and in both cases it is possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point.

When the aforementioned issues arise, rm-ANOVA requires exclusion of all subjects with missing observations from the analysis, thereby increasing costs for the study if the desired statistical power is not met with the remaining observations as it makes necessary to enroll more subjects [@ma2012]; and  raising the possibility that the rejection of those partial observations limits the demonstration of significant differences between groups. Additionally, rm-ANOVA uses a _post hoc_ analysis to assess the relevance between the measured response in different groups. This analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance. Because the p-value is highly variable [@halsey2015;@nuzzo2014], multiple comparisons can inflate the false positivity rate [@liu2010],consequently biasing the conclusions of the study.

During the last decade, the biomedical community has started to recognize the limitations of a rm-ANOVA approach in the analysis of longitudinal information. This is exemplified by the use of  linear mixed effects models (LMEMs)  in certain groups to analyze tumor longitudinal data [@skala2010;@vishwanath2009]. Briefly, these models incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g. the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population [@pinheiro2006].These models are more flexible than rm-ANOVA as they can accommodate missing observations for multiple subjects, and allow different modeling strategies for the variability within each measure in every subject [@west2014;@pinheiro2006].On the other hand, they impose restrictions in the distribution of the errors of the model and random effects [@gueorguieva2004;@schielzeth2020].

Additionally, another assumption made in both rm-ANOVA and LMEMs models is that a linear relationship exists between the observed response and the explanatory variables (covariates) across the study [@pinheiro2006].This common assumption to both rm-ANOVA and LMEMs causes the models to be restrictive in their inferences when used in longitudinal data that does not follow a linear trend. In biomedical research, this particular behavior in longitudinal has been reported in studies of tumor response to radio/chemotherapy in preclinical and clinical settings[@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994]. These studies have shown that that the collected signal does not follow a linear trend over time, and presents high variability at different time points, making the estimations of a LMEM or rm-ANOVA model inconsistent with the pattern of the observed variations. Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” results ( _p_<0.05) by using multiple comparisons between the model terms, this metric is inherently biased due to the discrepancy between the data and the model.

As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumed distribution) to determine the fit of the model while enabling inferences that are both adequate and consistent from a statistical perspective.

Generalized additive models (GAMs) are a subset of generalized linear models that use _smooth functions_ (hencefort _smooths_) to estimate the parameters of a model. They have been used in palaeolimnology, ecology and clinical studies to model longitudinal data [@woolway2016;@hefley2017;@ko2007]. Briefly, GAMs use a combination of multiple functions (basis functions) to construct the smooths of the model [@wood2017]. Their main advantage over LMEMs and rm-ANOVA is that the model specification is directed by the _smooths_ rather than by a parametric relationship. This allows a consistent fit of the model with the data, and estimations of significance using the terms of the model. However, certain assumptions about the data are necessary: a normal distribution and constant variance of the residuals with the mean response. Therefore, GAMs provide a more suitable statistical method to analyze biomedical longitudinal data, when these assumptions of the model are met by the data.

However, it is possible that the assumptions of GAMs for longitudinal data do not hold under  certain circumstances. In that case, the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information. Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine significant differences and are able to expand the number of comparisons and inferences derived form the analysis. On the other hand, the shift that Bayesian statistics represent from the traditional "frequentist" statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community. However,Bayesian theory is intuitive and shares some principles with "frequentist" statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017].

Additionally,the current development in computational tools, specifically the programming language $\textsf{R}$, enable a rapid implementation of GAMs and Bayesian models for longitudinal data.In particular, $\textsf{R}$ has an extensive collection of documentation, functions and libraries that speed up the initial stages of the analsys and that enable the use of complex statistical methods without requiring a specialized set of programming skills from the user.

Therefore, this study focuses in three areas in the analysis of longitudinal data from a biomedical perspective. First, it presents the limitations of (rm-ANOVA) and LMEMs over longitudinal data, and explains how these limitations in turn affect the results of the analysis. Secondly, it uses $\textsf{R}$ to simulate non-linear longitudinal data following previously reported values in the literature, and presents the implementation of GAMs as a statistical tool for longitudinal data. And finally, it introduces Bayesian statistics and presents their implementation with GAMs to demonstrate the differences and benefits of this approach. with an emphasis in reproducibility  by providing the code, simulated dataset and a step-by-step guide of the computational implementation of different models, this study aims to encourage the exploration of modern statistical methods for biomedical longitudinal data, and to improve the  statistical standards in biomedical research.

@@

#### 1.3 Bayesian brief introduction, and compare the results of 1.2 to the results of Bayesian

>>>>

A longitudinal study is defined as that which is designed to repeatedly measure a variable of interest in a group (or groups) of subjects. In biomedical research, this type of study is employed when the intention is to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study). Clinical studies of breast and neck cancer have used a longitudinal approach [@sio2016;@kamstra2015]. In the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT). Longitudinal studies have been also used to analyze tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].

Traditionally, a “frequentist” approach is used in biomedical research to derive inferences from a longitudinal study. Such statistical view derives its name from the fact that it regards probability as a limiting frequency[@wagenmakers2008] and its application is based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). This methodology makes two key assumptions regarding longitudinal data: constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as complete observations) [@gueorguieva2004;@schober2018].

However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases [@ugrinowitsch2004]. 
In the case of the number of observations during the study, different situations can arise that prevent the collection of complete measurements for all subjects:  in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and it is even possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point and therefore violating the _complete observations_ assumption of rm-ANOVA.

When incomplete observations occur, an rm-ANOVA requires the exclusion of all subjects with missing observations from the analysis. This can result in increased costs if the desired statistical power is not met with the remaining observations, as it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups. Additionally, rm-ANOVA uses a _post hoc_ analysis to assess differences between the measured response in different groups. A _post hoc_ analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance. Because the _p-value_ is highly variable, multiple comparisons can inflate the false positivity rate [@liu2010;@halsey2015], consequently biasing the conclusions of the study.

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information. This is exemplified by the  use of  linear mixed effects models (LMEMs)  in certain groups to analyze longitudinal data [@skala2010;@vishwanath2009]. Briefly, these models incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g. the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population. These models are more flexible than the traditional rm-ANOVA as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006]. On the other hand, they impose restrictions in the distribution of the errors  of the random effects [@gueorguieva2004].

One final assumption that is not initially evident for  both rm-ANOVA and LMEMs models pertains the relationship between the observed effect and the progression of the timeline of the study, which is expected to be linear [@pinheiro2006]. This common assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend. In biomedical research, a special case of this behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009]. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent with the observed variation. In other words, the in both cases the model tries to explain highly-variable data using a linear trend, consequently biasing the estimates. Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) by using multiple comparisons between the model terms, the use of this estimator is limited in such scenario because it is intended to work in models that have a reasonable agreement with the data.

As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, while enabling inferences that are both adequate and consistent from a statistical perspective.
In this regard, Generalized Additive Models (GAMs) present an alternative approach to analyze longitudinal data. While not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018]. There are several advantages of GAMs over LMEMs and rm-ANOVA models: They have an extensively developed theoretical background, enable the data to dictate the trend of the model, can accommodate missing observations and do not require constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more suitable statistical method to analyze non-linear biomedical longitudinal data.

 The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs. In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2011;@wood2016;@wood2017] that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g. hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. This programming language is also able to simulate data, an emerging strategy used to test statistical models [@haverkamp2017]. This allows to create and  explore different alternatives for analysis without collecting information in the field, and reduces the time window between experiment design and its implementation.

Therefore, the purpose of this study is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity and correlation structures. Secondly, the key theoretical elements of GAMS are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.  Thirdly, simulated data that follows the trend of previously reported values [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research and the implementation of GAMs in such scenario. Finally, reproducibility is emphasized by providing the code to generate the data and the implementation of GAMs in $\textsf{R}$, in conjunction with a step-by-step guide to fit models of increasing complexity.  
In summary, the exploration of modern statistical techniques to analyze longitudinal data may allow biomedical researchers to consider the use of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a linear trend, and will also help to improve the standards for reproducibility  in biomedical research.

@@

#### 1.3 Bayesian brief introduction, and compare the results of 1.2 to the results of Bayesianjo
